### 2015-08-31 Paper_v2.tex
Final Updates
Updated version of the paper-paper_v2

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses  
\usepackage{ifthen}  % Conditional 
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{comment}
\usepackage{lineno}

\urlstyle{rm}

@@

%\usepackage{geometry}
\usepackage{fullpage}

%\bibliographystyle{plain}
\bibliographystyle{msb}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large Modeling and Analysis of Hormone and Mitogenic Signal Integration in Prostate Cancer}}
\vspace{0.05in}
{\par \centering \large{ Katharine V. Rogers, Joseph A. Wayman, Ryan Tasseff, Caitlin Gee, Matthew P. DeLisa, 
and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering \large{School of Chemical and Biomolecular Engineering}}
{\par \centering \large{Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Modeling signal transduction in prostate cancer}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{Molecular Systems Biology}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853} 
{\par \centering Email: jdv27@cornell.edu} 
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Prostate cancer is the most common cancer in men and the second leading cause of cancer related death in the United States. 
Androgens, such as testosterone, are required for prostate cancer growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer. 
However, androgen ablation typically fails to permanently arrest cancer progression, often resulting in castration resistant prostate cancer (CRPC). 
CRPC is closely related to metastasis and decreased survival. 
In this study, we developed and analyzed a population of mathematical models describing growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
The model describes the integration of two simultaneous extracellular signaling inputs, namely, androgen and growth factors into a G1/S cell cycle checkpoint decision. 
Model parameters were identified from 43 studies in androgen dependent and resistant LNCaP cell lines. 
The model was validated by comparing simulations with an additional 29 data sets from LNCaP cell lines that were not used during training. 
Additionally, data from four drug trials was also used to evaluate the model’s performance.
Sensitivity analysis, conducted over an ensemble of prostate signaling models, suggested that in an androgen free environment general translation and transcription was more sensitive in androgen dependent cells, while in androgen independent cells the PI3K and MAPK pathway species were more sensitive. 
In a constant DHT environment sensitive species were conserved between the cell lines. 
These results suggest targeting the PI3K and MAPK pathways in addition to anti-androgen therapies as a treatment for CRPC.

\vspace{0.5in}
{\noindent \textbf{Keywords:}~Prostate cancer, signal transduction, mathematical modeling}

@@

%
%Sipuleucel-T is a first generation cancer vaccine \cite{Kantoff2010}, while radium-223 is a targeted alpha emitter for bone metastasis \cite{Parker2013}. 
%Abiraterone is an inhibitor of cytochrome P450 17A1 (CYP17), a critical enzyme in androgen biosynthesis \cite{Sartor2013} while enzalutamide inhibits androgen receptor (AR) translocation to the nucleus, as well as DNA binding and coactivator recruitment \cite{Scher2012}.
%Thus, understanding the operation of the biological pathways which ultimately facilitate the transition to CRPC is vital to the discovery of new therapeutic strategies against this disease.
%The model focuses on the outlaw pathway, in which constitutively or overly-activated receptor tyrosine kinases (RTKs) stimulate AKT and mitogen-activated protein kinases (MAPKs), which in turn activate cytosolic androgen receptor (AR) in the absence of androgen. Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation. Thus, the outlaw pathway encodes crosstalk between RTK-dependent kinase and AR activation inside of a positive feedback loop. The model was validated with multiple experiments in LNCaP cell lines. In addition  For example, in response to enzalutamide, an AR inhibitor, the model predicted a decline in PSA expression similar to patient response.
%and there is limited data on combination therapies or the use of biomarkers to develop personalized treatment strategies.

Prostate cancer (PCa) is the most commonly diagnosed cancer and the second leading cause of cancer-related death in men in the United States \cite{Siegel2013}. 
Initially, PCa cells depend upon the activation of cytosolic androgen receptors (AR) by androgen hormones, such as testosterone, for survival and growth. 
Androgen ablation in combination with radiation or chemotherapy remains the primary non-surgical treatment for androgen dependent prostate cancer (ADPC) \cite{Huggins1967}. 
However, androgen ablation typically fails to permanently arrest cancer progression as malfunctioning cells eventually lose androgen sensitivity and proliferate without hormone. 
The loss of androgen sensitivity results in castration resistant prostate cancer (CRPC), a phenotype closely linked with metastasis and greatly reduced survival \cite{Harris2009}.
Currently, there are six approved treatments that demonstrate a survival advantage in patients with metastatic CRPC, each of these target diverse aspects of the disease \cite{Sartor2013}.
The taxane family members docetaxel and cabazitaxel interact with microtubule stability \cite{Tannock2004,Bono2010}, while abiraterone \cite{Sartor2013} or enzalutamide \cite{Scher2012} interfere with androgen signaling by blocking androgen formation or nuclear translocation, respectively.
Other approved treatments are non-specific to PCa. 
For example, general treatments such as sipuleucel-T, a first generation cancer vaccine \cite{Kantoff2010}, 
and radium-223, an alpha emitter which targets bone metastasis \cite{Parker2013}, are both approved to treat CRPC.
Unfortunately, regardless of the therapeutic approach, the survival advantage of these treatments is typically only a few months. 
Thus, understanding the molecular basis of the loss of androgen sensitivity in CRPC could be an important step for the development of the next generation of therapies with a prolonged survival
advantage.

Androgen-induced proliferation and survival depends upon many coordinated signal transduction and gene expression events. 
Androgen Receptor (AR) is part of the nuclear hormone receptor superfamily, which includes other important cancer targets such as progesterone receptor (PR) and estrogen receptor (ER) in breast cancer \cite{Aranda2001}. 
Nuclear hormone receptors act as ligand dependent transcription factors interacting with specific DNA sequences of target genes as either monomers, heterodimers, or homodimers; AR, PR, and ER act as homodimers. 
In the case of AR these specific DNA sequences are known as androgen response elements (ARE) \cite{Mangelsdorf1995}.
In the absence of androgen, AR is predominately found in the cytoplasm bound to heat shock proteins (HSP) \cite{Prescott2006}.
%Nuclear Hormone Receptors (Reviewed \cite{Aranda2001}) and AR in disease reviewed in \cite{Matsumoto2013}
Androgen, either testosterone or testosterone metabolites such as 5$\alpha$-dihydrotestosterone (DHT), enter prostate cells and interact with the cytosolic androgen receptor (AR). 
The interaction of DHT with AR promotes the dissociation of AR from chaperones such as HSP \citep{Pratt:1997kx} and its subsequent dimerization, phosphorylation and translocation to the nucleus (reviewed by Brinkmann \textit{et al.} \citep{Brinkmann1999}). 
Activated nuclear AR drives a gene expression program broadly referred to as androgen action, that promotes both proliferation and survival.
In addition to many genes including itself, activated nuclear AR promotes the expression and secretion of prostate specific antigen (PSA), arguably the best known PCa biomarker \citep{Feldman2001}.
PSA is commonly used as a prostate cancer indicator, although its prognostic ability is controversial \citep{Attard2009,Hoffman2011,Moyer2012}. 
In CRPC, AR signals in the absence of androgens. 
Androgen dependent (AD) prostate cells can become castration resistant (CR) through several possible mechanisms, including constitutively amplified AR expression and altered AR sensitivity to testosterone or other non-androgenic molecules \citep{Feldman2001}. 
In this study, we focused on the aberrant activation of AR by kinase signaling cascades, sometimes called the outlaw pathway. 
Outlaw pathway activation is driven by over-activated receptor tyrosine kinases (RTKs), a common pathology in many cancer types including PCa \citep{Slamon1989,Craft1999}. 
RTKs stimulate downstream kinases, including the AKT and mitogen-activated protein kinase (MAPK) pathways, which promote AR phosphorylation and dimerization in the absence of an androgen signal \citep{Craft1999,Yeh1999}. 
Interestingly, among the few genes activated AR represses is cellular prostatic acid phosphatase (cPAcP), itself a key regulatory of RTK activation \cite{Veeramani2005}. 
Thus, in CRPC the androgen program is initiated without the corresponding extracellular hormone cue, potentially from crosstalk between growth factor and hormone receptor pathways.

In this study, we developed a mathematical model of growth factor and hormone signal integration in androgen dependent, intermediate and resistant prostate cancer cells. 
We used this model to better understand which components and processes were differentially important in AD versus CR cells.
The new model architecture was a significant advance over our previous prostate signaling model \citep{Tasseff2010}. 
We added the regulated expression of ten additional proteins, including the cell cycle restriction point proteins cyclin D (and the differentially spliced variants cyclin D1a and cyclin D1b), cyclin E, cyclin-dependent kinase inhibitor 1A (p21Cip1), and cyclin-dependent kinase inhibitor 1B (p27Kip1). 
Also, we included the Rb/E2F pathway, expanded our description of the activation of the mammalian target of rapamycin (mTOR) protein and its role in translation initiation, and included the regulation of AR action by cyclin D1a and E2F. 
However, this upgraded architecture, while increasing the biological scope of the model, also expanded the number of unknown model parameters. 
To estimate these parameters, we used multiobjective optimization in combination with dynamic and steady-state data sets generated in AD, intermediate and CR LNCaP cell lines. 
We identified a population of approximately N = 5000 models (from well over a million candidate models) which described both AD and CR data sets using a single model structure.
Furthermore, we tested the model using an additional 29 LNCaP data sets not used for model training, along with four drug studies.
We analyzed the model population using sensitivity and robustness analysis to uncover differentially important mechanisms in AD versus CR cell lines. 
In the presence of androgen, the sensitivity profile was similar between AD and CR cells. 
Components of the MAPK and PI3K pathways were highly fragile, irrespective of the level of androgen dependence. 
However, in the absence of androgen, there were 609 statistically significant shifts in species sensitivity between AD and CR cells.
Of these, 108 were larger than one standard deviation above the mean.
In CR cells, HER2 activation of the MAPK and PI3K pathways was significantly more important, as was AR activation through the MAPK pathway. 
On the other hand, components of the translation and transcription infrastructure were differentially more important in AD cells in the absence of androgen.  
Taken together, our analysis suggested that independently targeting the PI3K or MAPK pathways in combination with anti-androgen therapies could perhaps be an effective treatment strategy for CRPC.

%(323 more and 287 less sensitive)
%In this study, we rank-ordered the importance of the molecular components controlling the transition from ADPC to CRPC.
%Toward this, we developed and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes governing the 
%transition from androgen dependent to castration resistant prostate cancer.
%The objective of this study was to rank-order the importance of components in CRPC and ADPC using computational tools. 
%We formulated and analyzed an ensemble of mechanistic mathematical models of the signal transduction processes in CRPC and ADPC 
%that was an extension of our previous work .
%In particular, we focused on the outlaw pathway and the possible crosstalk between RTK-dependent kinases and AR activation. 
%First, we expanded the original model by adding ten  
%Third, we added additional co-activators, such as cyclin E, and inhibitors of AR, such as
%ODE models typically contain a large number of unknown model parameters, which must be estimated  
%Due to uncertainty in parameter estimation and to account for differences between individual cells and cell sublines, an ensemble of parameter sets was used to model the system. 
%Proteins that were not modeled at a gene regulatory level were generated using zeroth order reactions. 
%All protein and RNA species were degraded with first order reaction rates. 
%To obtain an ensemble of parameter sets, we used multiobjective optimization. 
%Relative concentration values were quantified 
%Training data came from experiments on LNCaP sublines. 
%We used , previously developed in the lab \cite{Song2010}, to determine a diverse ensemble of models. 
%C-33 cells, having undergone fewer than 33 passages, are considered AD while C-51 cells, having between 34 and 80 passages, are considered an intermediate prostate cancer cell line. 
%C-81 (passage numbers beyond 80) display a CR phenotype \cite{Lin2000}. 
%http://endo.endojournals.org/content/140/11/5054.long or http://endo.endojournals.org/content/140/11/5054.long#ref-19

>>>>

\documentclass[12pt]{article}
% Load packages
\usepackage{url}  % Formatting web addresses
\usepackage{ifthen}  % Conditional
\usepackage{multicol}   %Columns
\usepackage[utf8]{inputenc} %unicode support
\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{epsfig}
\usepackage{epstopdf}
\usepackage{graphicx}
\usepackage[margin=0.1pt,font=footnotesize,labelfont=bf]{caption}
\usepackage{setspace}
%\usepackage{longtable}
\usepackage{colortbl}
%\usepackage{palatino,lettrine}
%\usepackage{times}
%\usepackage[applemac]{inputenc} %applemac support if unicode package fails
%\usepackage[latin1]{inputenc} %UNIX support if unicode package fails
\usepackage[wide]{sidecap}
%\usepackage[authoryear,round,comma,sort&compress]{natbib}
\usepackage[square,sort,comma,numbers,sort&compress]{natbib}
%\usepackage[authoryear,round]{natbib}
\usepackage{supertabular}
\usepackage{simplemargins}
\usepackage{fullpage}
\usepackage{comment}
\usepackage{lineno}
%\usepackage{chicago}
\usepackage{textcomp}

%\usepackage[space]{cite}
\urlstyle{rm}

@@

%\bibliographystyle{chicago}
\bibliographystyle{plos2009}

@@

% Begin ...
\begin{document}
\begin{titlepage}
{\par\centering\textbf{\Large {Dynamically dimensioned particle swarm optimization (DdPSO): A meta-heuristic for parameter estimation in biochemical models}}}
\vspace{0.05in}
{\par \centering \large{Adithya Sagar, Christine Shoemaker$^{\dag}$ and Jeffrey D. Varner$^{*}$}}
\vspace{0.10in}
{\par \centering {School of Chemical and Biomolecular Engineering}}
{\par \centering {$^{\dag}$School of Civil and Environmental Engineering}}
{\par \centering {Cornell University, Ithaca NY 14853}}
\vspace{0.1in}
{\par \centering \textbf{Running Title:}~Parameter estimation in biochemical models}
\vspace{0.1in}
{\par \centering \textbf{To be submitted:}~\emph{PLoS ONE}}
\vspace{0.5in}
{\par \centering $^{*}$Corresponding author:}
{\par \centering Jeffrey D. Varner,}
{\par \centering Associate Professor, School of Chemical and Biomolecular Engineering,}
{\par \centering 244 Olin Hall, Cornell University, Ithaca NY, 14853}
{\par \centering Email: jdv27@cornell.edu}
{\par \centering Phone: (607) 255 - 4258}
{\par \centering Fax: (607) 255 - 9166}
\end{titlepage}
\date{}
\thispagestyle{empty}
\pagebreak
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section*{Abstract}
Mathematical modeling is as a powerful tool to analyze, and ultimately design biochemical networks.
However, estimation of biochemical model parameters can be a significant challenge.
Parameter estimation in biochemical models generally involves expensive function evaluations, making it difficult to quickly obtain optimal solutions.
Additionally, biochemical models often have many local minima or maxima which further complicates the parameter estimation problem.
In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamically dimensioned search. DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamically dimensioned search. We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), on published biochemical parameter estimation benchmark problems and
lastly on a model of the human coagulation cascade. For each test we performed 25 trials with 4000 function evaluations per trial, and compared the performance of DdPSO with other commonly
used meta-heuristic approaches such as differential evolution (DE), particle swarm optimization (PSO), simulated annealing (SA) and dynamically dimensioned search (DDS).
For the wide range of problems that we considered, DdPSO outperformed the other meta-heuristic approaches despite a limited number of function evaluations.
We further tested the predictive power of the coagulation model parameters against data not used in training, and found good agreement between model simulations and experimental measurements.
Taken together, DdPSO is a promising meta-heuristic approach for the estimation of biochemical model parameters in relatively few function evaluations.

{\noindent \textbf{Keywords:}~Parameter identification, Meta-heuristic optimization, Biochemical modeling}

% Extra abstract
% Mathematical modeling of biological systems with multiple feed back loops is one such area where parameter estimation is a difficult non-linear optimization problem. This difficulty is further compounded when dealing with parameter vectors of high dimensions.

%In this study, we present the dynamically dimensioned particle
%a novel meta-heuristic approach that combines a variant of particle swarm optimization (PSO) along with dynamically dimensioned search (DDS) to obtain near optimal solutions of high dimensional biochemical networks within a relatively few function evaluations.
%We use a particle swarm optimization technique that uses multi-swarms to generate candidate vectors which are then greedily updated using DDS by dynamically varying the perturbed parameter dimensions. We tested this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and also against DDS alone. The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on commonly used test functions (Ackley and Rosenbrock) for global optimization and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.

@@

Cells process nutrients and respond to changes in their environment using complex biochemical networks. These networks contain thousands of components
interconnected through nonlinear enzyme catalyzed reactions. Mathematical modeling has evolved as a powerful paradigm to analyze, and ultimately design these complex networks \cite{assmus2006dynamics, Riel:2006aa,Jaqaman:2006aa,kitano2002systems,hood2004systems}. Mathematical modeling of biochemical networks is often an iterative process.
First, models are formulated from biochemical knowledge, and then model parameters are estimated using experimental data \cite{Aldridge:2006aa,banga2008optimization,ashyraliyev2009systems}.
Parameter estimation is typically framed as a non-linear optimization problem wherein the residual (or objective function) between experimental data and model simulations is minimized using an optimization strategy \cite{moles2003parameter}. The optimal parameters obtained from training are then used to validate the model on hitherto unseen experimental data. If validation fails, model construction and calibration are repeated iteratively until satisfactory results are obtained.

Parameter estimation is a major challenge in the development of large biochemical models.
Although parameter estimation has been a well studied problem in engineering for decades \cite{nieman1971review,beck1977parameter,young1981parameter,beck1998inverse},
the complex dynamics of large biological systems and noisy, often incomplete experimental data pose a unique estimation challenge.
Most of the optimization problems pertaining to biological systems are non-linear and multi-modal i.e. most of them have multiple local minima or maxima \cite{moles2003parameter,banga2008optimization}. The non-linearity of the problem coupled with multi-modality generally renders the local optimization techniques like pattern search method \cite{hooke1961direct}, Nelder-Mead simplex method \cite{nelder1965simplex}, steepest descent method or Levenberg-Marquardt method \cite{more1978levenberg} incapable of reliably obtaining an optimal solution since they generally stop at the first local minimum. Though deterministic global optimization techniques (for example algorithms based on branch and bound framework) can handle non-linearity and multi-modality \cite{esposito2000deterministic,horst2013global}, the absence of derivative information, discontinuity of the objective functions, non-smooth regions or the lack of any knowledge about the objective function severely hamper the use of these techniques.

Meta-heuristic stochastic optimization approaches like Genetic Algorithms (GAs), Simulated Annealing (SA) \cite{kirkpatrick1983optimization}, Evolutionary Programming and population based searches like Differential Evolution (DE) \cite{storn1997differential} have shown promise in this regard \cite{sun2012parameter}. They do not make any assumptions about the structure of objective function or require any \textit{a priori} information about the objective function. Though they do not guarantee strong convergence, these approaches are effective in finding near optimal solutions.  Mendes et al. \cite{mendes1998non} used Simulated Annealing to estimate rate constants for the irreversible inhibition of HIV proteinase, Modchang et al. \cite {modchang2008mathematical} used Genetic Algorithms to estimate parameters for a mathematical model of signal transduction, Differential Evolution based approaches have been an effective approach on various systems \cite{tsai2005evolutionary,wang2001hybrid,noman2007inferring}. Tashkova et al. \cite{tashkova2011parameter} compared different meta-heuristics for parameter estimation on a dynamic model of endocytosis and showed that DE was the most effective. Banga and co-workers have applied scatter-search based methods \cite{villaverde2012cooperative,rodriguez2006novel,egea2007scatter} to estimate parameters on non-linear biological processes. Hybrid approaches that combine a meta-heuristic with a local optimization search, wherein a near globally optimal solution that is obtained with a meta-heuristic is further refined using a local search have also become popular. Villaverde et al. \cite{villaverde2015biopredyn} combined scatter search with local search methods for parameter estimation in large scale systems biology models. Fan et al. recently showed that population based meta-heuristics along with decomposition based methods can be used to model gene circuits from mRNA data \cite{fan2015parameter}.Despite these successes, a  major drawback with most of these metaheuristic approaches is the vast number of objective function evaluations they take in arriving at good search regions. As the models grow in size and complexity and the number of dimensions of parameter vector increases, evaluation of the objective function becomes computationally expensive. Thus performing a large number of evaluations is not computationally feasible.

In many of these high dimensional problems approaching an exact solution may not be necessary. Gutenkust et al. \cite{gutenkunst2007universally} showed that a number of systems biology models are 'sloppy'. Sloppy systems have specific parameter combinations that largely define the dynamics of the system. Large perturbations to the rest of the parameters does not greatly impact the system dynamics. Ensemble approaches \cite{song2010ensembles,luan2010ensembles} have exploited this aspect to describe the dynamics of biological systems including coagulation which can be described using only a set of key species or parameters \cite{sagar2015dynamic}. Tolson and Shoemaker \cite{tolson2007dynamically} showed through Dynamically Dimensioned Search (DDS) that high-dimensional watershed models can be calibrated quickly by perturbing only a subset of dimensions.

In this study, we developed dynamically dimensioned particle swarm optimization (DdPSO), a novel meta-heurstic that combined the global search capability of multi-swarm particle swarm optimization with dynamic dimensioned search. The objective of DdPSO was to obtain near optimal parameter estimates for large biochemical models within a relatively few function evaluations.
DdPSO uses a multi-swarm particle swarm optimization technique to generate candidate solution vectors which are then greedily updated using dynamic dimensioned search.
We tested the performance of DdPSO on commonly used test functions for global optimization (Ackley and Rosenbrock), published biochemical parameter estimation benchmark problems and
lastly on a model of the human coagulation cascade. DdPSO outperformed common meta-heuristic approaches like Differential Evolution (DE), Particle Swarm Optimization (PSO), Simulated Annealing (SA) and dynamically dimensioned search (DDS) on the benchmark and coagulation problems.

% Extra text from the introduction -
%With the capacity to construct very large models using various mathematical formalisms \cite{chen2009input,tasseff2011modeling,luan2007computationally,mo2007genome,orth2011comprehensive,karr2012whole,buchel2013path2models,smallbone2010towards}, more often than not,

%this algorithm (25 trials with 4000 function evaluations in each trial) on a biochemical network of coagulation (148 parameters and 92 species) and compared it's performance against other meta-heuristics  The new algorithm outperforms all the other meta-heuristics on the coagulation model. The parameter vectors obtained using this approach fit the experimental data well and also make accurate enough predictions on unseen experimental data. We also performed this comparison on  and found the same behavior. Further we used two recently published benchmark problems, a genome wide kinetic model with 1759 parameters and a metabolic model of Chinese Hamster Ovary cells with 117 parameters to evaluate the performance of our approach. We  surprisingly performed well on these benchmarks and obtained the nominal parameter vector with just 4000 function evaluations in both cases.
%took into cognizance the fact that it may not be necessary to obtain the exact solution for high dimensional biological systems and that good enough solutions can be quickly obtained without expending a lot of objective function evaluations. Hence our current approach uses the power of population based heuristics along with DDS to obtain near optimal or good solutions. Though in theory we can combine any meta-heuristic with DDS, for the current purpose we used Particle Swarm Optimization (PSO). PSO unlike Genetic Algorithm (GA) or Differential Evolution (DE) does not have complex operations like cross over, mutation or recombination. It is simple to use and does not have a lot of parameters associated with other heuristics. However PSO is known to rapidly converge to a local optimum and thus several variants \cite{peer2003using,zhan2009adaptive,li2007fast} have been developed of which Multi swarm particle swarm optimization approaches (MLSPSO) \cite{zhao2008dynamic,liang2005dynamic} are one of the effective ones. We used MLSPSO to preclude bad regions of search and thereafter search within these regions using DDS.

\clearpage
\section*{Results}
\subsection*{Dynamically dimensioned particle swarm optimization (DdPSO).}
DdPSO is a novel meta-heuristic that combines multi-swarm based particle swarm methods with dynamically dimensioned search (Fig. \ref{fig-algorithm}).
The goal of this approach is to obtain optimal or near optimal parameters for high-dimensional complex biological systems within a pre-specified number of function evaluations. We randomly initialized a swarm of $\mathcal{K}$-dimensional particles (represented as ${x}_{i}$), wherein each of these particles corresponds to a $\mathcal{K}$-dimensional parameter vector. After initialization, the particles were grouped into different sub-swarms randomly. Thereafter within each sub-swarm ${S}_{k}$,  particles were updated according to the following rule.
\begin{eqnarray}
	\mathbf{x}_{i,j} &=&\theta_{1,j-1}\mathbf{x}_{i,j-1} + \theta_{2}\mathbf{r}_{1}\left(\mathcal{L}_{i} - \mathbf{x}_{i,j-1}\right) + \theta_{3}\mathbf{r}_{2}\left(\mathcal{GL}_{k} - \mathbf{x}_{i,j-1}\right)
\end{eqnarray}
where $\left(\theta_{1},\theta_{2},\theta_{3}\right)$ are adjustable parameters, $\mathcal{L}_{i}$ denotes the best local solution found by particle $i$ till function evaluation $j$-$1$, and
$\mathcal{GL}_{k}$ denotes the best local solution found over the population of all particles within the swarm ${S}_{k}$. The quantities $r_{1}$ and $r_{2}$ denote uniform random vectors with the same dimension as the number of unknown model
parameters ($\mathcal{K}\times{1}$). In our algorithm the parameter $\theta_{1,j-1}$ depends on the function evaluations and is controlled according to the following equation
\begin{eqnarray}
	\mathbf \theta_{1,j}&=&((\mathbf{N}-{j})*(\mathbf{w}_{max}-\mathbf{w}_{min}))/(\mathbf{N}-{1}) + \mathbf{w}_{min}
\end{eqnarray}
where $\mathbf{N}$ represents the total number of function evaluations, $\mathbf{w}_{max}$ and $\mathbf{w}_{min}$ are the maximum and minimum inertia weights respectively. While updating the particle, we made sure all dimensions of the solution represented by the particle were within bounds using a reflection boundary condition. After every $\mathbf{g}$  function evaluations, the particles within all sub-swarms were mixed and then randomly redistributed to a new sub-swarm. The particles were then again updated according Eq.1. This process continued till $\mathcal{FR}*\mathbf{N}$ number of functions evaluations, where $\mathcal{FR}$ represents the fraction of evaluations with the multi-swarms. At the end of these function evaluations, we froze all the solutions represented by various particles and chose the particle with best solution among $\mathcal{GL}_{1} \cdots \mathcal{GL}_{NS}$ as the initial candidate vector $\mathcal{G}$ for the remaining $({1}-\mathcal{FR})*\mathbf{N}$ number of function evaluations.

This particle was then updated according to the following rule
\begin{eqnarray}
	\mathcal{G}_{new}(\mathbf{J})& = &\mathcal{G}(\mathbf{J})+\mathbf{r}_{normal}(\mathbf{J})\sigma(\mathbf{J})
\end{eqnarray}
where $\mathbf{J}$ represents the set containing the specific dimensions being perturbed, ${r}_{normal}$ denotes a normal random vector of the same dimensions as $\mathcal{G}$. $\sigma$ is the amplitude of perturbation given by following equation:
\begin{eqnarray}
	\sigma& = &\mathbf{R}(\mathcal{MAX} -\mathcal{MIN})
\end{eqnarray}
where $\mathbf{R}$ is the scalar perturbation size parameter, $\mathcal{MAX}$ and $\mathcal{MIN}$ are ($\mathcal{K}\times{1}$) vectors that represent the maximum and minimum bounds on each dimension. The set $\mathbf{J}$ was constructed using a probability function $\mathcal{P}_{j}$ that represented the probability whether a specific dimension $j$ was perturbed or not.  This function is a monotonically decreasing function that decreases with the number of function evaluations. $\mathcal{P}_{j}$ can be any monotonically decreasing function, in our approach we used the following function:
\begin{eqnarray}
	\mathcal{P}_{j}&=&{1}-\log(j/ (\beta({1}-\mathcal{FR})*\mathbf{N}))
\end{eqnarray}
where $\beta$ is the perturbation frequency probability modulator. Thus the number of dimensions of the candidate vector that are updated or perturbed decreases with the as the number of function evaluations increase. These updates are greedy in nature, so $\mathcal{G}_{new}$ becomes the new solution vector only if it is better than the old one $\mathcal{G}$.

\subsection*{Performance of DdPSO on benchmark problems.}
We compared the performance of DdPSO with algorithms on commonly used test functions for global optimization, as well as benchmark biochemical problems from literature (Table \ref{table:benchmark-problems}). We used a 300 dimensional Rastrigin function and 300 dimensional Ackley function. Both Ackley and Rastrigin have multiple local minima and maxima and attain a global minimum value of 0. We tested the performance of the 5 different heuristic approaches on these two functions. In each experiment we again fixed the number of function evaluations at 4000 and ran 25 experiments. In both cases (Figure S6 A and B) we see that the error convergence rate for DDSMLSPSO is much faster as compared to the other four s and it finds the global minimum of 0 in both cases.

Villaverde and co-workers recently published a set of benchmark problems to evaluate parameter estimation methods \cite{villaverde2015biopredyn}. From a computational cost perspective problems they categorized the problems as most expensive, intermediate and least expensive. We evaluated the performance of our algorithms on a problem from the most expensive and least expensive categories. The first problem is a genome wide kinetic model of \textit{Saccharomyces cerevisiae} with 1759 parameters (henceforth referred to as problem B1).  The second problem (henceforth referred to as problem B4) is a metabolic model of Chinese Hamster Ovary (CHO) with 35 metabolites, 32 reactions and 117 parameters. In both cases pseudo time series data was generated. For problem B1, the time series data consisted of 44 observables and for problem B4 the data corresponded to 13 different metabolites. We fixed the number of function evaluations at 4000 for DDSMLSPSO and trained both the models against the pseudo experimental data. We recaptured the 'nominal' parameters (Figure S7 A and B)  in both cases within 4000 evaluations. The final objective function value (Table 1) is an order smaller for B1 possibly due to overfitting. In Table 1 we summarize the results obtained on the coagulation model and the benchmark problems.

\subsection*{Performance of DdPSO on a model of the human coagulation cascade.}
We compared the performance of DdPSO on a model of blood coagulation against commonly used meta-heuristics like simulated annealing (SA), differential evolution (DE), particle swarm optimization (PSO),and dynamically dimensioned search (DDS). Coagulation is an archetype biochemical network that is highly interconnected and tightly regulated with multiple positive and negative feedback loops (Fig. \ref{fig-coagulation-network}). The biochemistry underlying coagulation, though quite complex has been well studied \cite{mann2003dynamics,mann2003all,mann2003thrombin,vogler2009contact,diamond2013systems,fogelson2005coagulation,anand2003model}, and reliable experimental coagulation models have been developed \cite{hockin2002model,chatterjee2010systems,mann2006models,luan2007computationally}. This makes it an ideal system for mathematical modeling and parameter estimation. Coagulation is regulated by a set of serine proteases also known as coagulation factors and blood platelets. The coagulation factors are generally in an inactive state and are known as zymogens. These zymogens are activated through certain triggers. Trigger events like injury or trauma or sepsis expose factors like collagen, tissue factor and von Willebrand factor (vWF) to blood. The exposure of these factors to blood kick-starts a series of convergent cascades that lead to conversion of zymogen prothrombinase to thrombin.   Luan et al. modeled coagulation using coupled non-linear ordinary differential equations with 148 reactions and 92 species \cite{luan2007computationally} and validated model using 21 published datasets.
